Co-operation between the sexes trumps selfish control of female meiotic drive. Or Sperm do not evolve to collaborate in female meiotic drive. Or Why does female meiosis allow the opportunity for exploitation by selfish sperm?

### 2014-02-14 writeup_and_notes/male_dependent_drive.tex
adding to appendix

\begin{figure}
	\rotatebox{90}{\includegraphics[width = 0.8
          \textwidth]{Figures/Sperm_egg_cartoon_paper_Fig2.eps}}
\caption{Dynamics in 3 allele system
\bf{A)} The sperm-dependent transmission probabilities of alleles through
female meiosis (female germ cells depicted as circles, sperm indicated by wiggly
tails). In each case the top and bottom entry in each cell give the
transmisson probability of the left and right allele in the female
being transmitted to the egg. In the left panel we show the transmission probabilities for
  A, B, and B'. Where B' is a driver allele which is either promotes
  itself through through female meiosis when sperm (a B$^+$ allele,
  $d'>d$) or self-avoiding through the interaction of sperm with female meiosis (a B$^-$ allele,
$d'<d$). In the right panel we depict the transmission probabilities for
  A, B, and A$^-$ allele a non-driven allele that acts to reduce the
  effects of female drive through sperm.
\bf{B) Frequency trajectories in the three allele system. }
}  \label{Eggsperm_3_allele_cartoon}
\end{figure}

@@

\yb{??In the SUPP we explore the dynamics of a sperm modifier of female drive unlinked to the female driver??}

@@

\section*{Appendix}
In our first scenerio allele $A$ is the ancestral allele and allele
$B$ is our sperm dependent drive allele.
The sperm dependence of the drive is a form of assortative
transmission (mating), and so we have to keep track of the genotype frequencies.
Our genotype frequencies at birth are $x_{AA},~x_{AB},$ and
$x_{BB}$. These three genotypes have fitnesses $w_{AA},~w_{AB},$ and
$w_{BB}$, which are identical between males and females. As the
fitnesses are the same in males and females the frequencies of the
genotypes are identical in males and females, but in the following
equations they are marked with $\Mars$ and $\Venus$ respectively to
allow the reader to more easily follow the logic.
The frequency of the three genotypes at birth in the next generation is:
\begin{equation}
x'_{BB} =\frac{1}{\bar{w} } \Big( w_{BB} x_{BB}^{\Venus} (w_{BB} x_{BB}^{\Mars} +\frac{1}{2}
  w_{AB} x_{AB}^{\Mars} ) + w_{AB} x_{AB}^{\Venus} (d w_{BB}
  x_{BB}^{\Mars} +\frac{1}{2} d
w_{AB} x_{AB}^{\Mars} ) \Big),  \label{eqn.A1}
\end{equation}
\begin{equation}
x'_{AA} =\frac{1}{\bar{w} } \Big(  w_{AA} x_{AA}^{\Venus} (w_{AA} x_{AA}^{\Mars} + \frac{1}{2}
   w_{AB} x_{AB}^{\Mars}) + w_{AB} x_{AB}^{\Venus}(\frac{1}{2} w_{AA} x_{AA}^{\Mars}+ \frac{1}{2} 
  w_{AB} x_{AB}^{\Mars}) \Big),  \label{eqn.A2}
\end{equation}
and
\begin{eqnarray}
x'_{AB} = & \frac{1}{\bar{w} } \Big( w_{AA} x_{AA}^{\Venus}  ( w_{BB}  x_{BB}^{\Mars} + \frac{1}{2}
  w_{AB}  x_{AB}^{\Mars} ) +w_{BB} x_{BB}^{\Venus} ( w_{AA}  x_{AA}^{\Mars} + \frac{1}{2}
  w_{AB}  x_{AB}^{\Mars} )  \nonumber \\
& + x_{AB}^{\Venus}(\frac{1}{2}  w_{AA}  x_{AA}^{\Mars}+\frac{1}{2} (1-d)
 w_{AB}  x_{AB}^{\Mars} + \frac{1}{4}  w_{AB}
 x_{AB}^{\Mars} + (1-d)  w_{BB}  x_{BB}^{\Mars} ) \Big) \label{eqn.A3}
\end{eqnarray}
where the mean fitness of the population is
\begin{equation}
\bar{w} = w_{AA} x_{AA}+w_{AB}  x_{AB} +w_{BB}  x_{BB}.
\end{equation}

\bibliography{refs}

@@

\begin{figure}
\includegraphics[width = 0.8 \textwidth]{Figures/bistable_x_vs_d_additive_s.eps} 
\caption{Bistable green-bearded allele when selection cost (s) acts
  additively, for an allele whose effect
 on female meoisis is mediated by sperm allele. }  \label{bistable_additive}
\end{figure}

\begin{figure}
\includegraphics[width = 0.8 \textwidth]{Figures/effect_of_dominance_on_invasion_space_one_graph.eps} 
\caption{Invasion analysis of green-bearded sperm allele whose effect
 on female meoisis is mediated by the genotype of the fertilizing
 male. The area below the solid/dotted line is the parameter region
 where the allele can invade/fix (from $1/1000$ and $999/1000$
 \gc{maybe should lower this}). Note
 that all of the fixation lines are on top of each other as dominance
 parameter makes little difference here. The allele will be
 polymorphic in the parameter space where the solid (invasion) line
 falls above the dotted (fixation) line. Note that this is a very
 narrow parameter range, and only exists if the allele has a
 reasonable degree on dominance}  \label{Effect_of_dominance}
\end{figure}

>>>>

\begin{figure}
	\rotatebox{90}{\includegraphics[width = 0.8
          \textwidth]{Figures/Sperm_egg_cartoon_paper_Fig2.eps}}
\caption{Dynamics in 3 allele system
{\bf (A) } The sperm-dependent transmission probabilities of alleles through
female meiosis (female germ cells depicted as circles, sperm indicated by wiggly
tails). In each case the top and bottom entry in each cell give the
transmisson probability of the left and right allele in the female
being transmitted to the egg. In the left panel we show the transmission probabilities for
  A, B, and B'. Where B' is a driver allele which is either promotes
  itself through through female meiosis when sperm (a B$^+$ allele,
  $d'>d$) or self-avoiding through the interaction of sperm with female meiosis (a B$^-$ allele,
$d'<d$). In the right panel we depict the transmission probabilities for
  A, B, and A$^-$ allele a non-driven allele that acts to reduce the
  effects of female drive through sperm.
{\bf (B)} Frequency trajectories in the three allele system. }
 \label{Eggsperm_3_allele_cartoon}
\end{figure}

@@

%\yb{??In the SUPP we explore the dynamics of a sperm modifier of female drive unlinked to the female driver??}

@@

\section*{Appendix}
In our first scenerio allele $A$ is the ancestral allele and allele
$B$ is our sperm dependent drive allele.
The sperm dependence of the drive is a form of assortative
transmission (mating), and so we have to keep track of the genotype frequencies.
Our genotype frequencies at birth are $x_{AA},~x_{AB},$ and
$x_{BB}$. These three genotypes have fitnesses $w_{AA},~w_{AB},$ and
$w_{BB}$, which are identical between males and females. As the
fitnesses are the same in males and females the frequencies of the
genotypes are identical in males and females, but in the following
equations they are marked with $\Mars$ and $\Venus$ respectively to
allow the reader to more easily follow the logic.
The frequency of the three genotypes at birth in the next generation is:
\begin{equation}
x'_{BB} =\frac{1}{\bar{w} } \Big( w_{BB} x_{BB}^{\Venus} (w_{BB} x_{BB}^{\Mars} +\frac{1}{2}
  w_{AB} x_{AB}^{\Mars} ) + w_{AB} x_{AB}^{\Venus} (d w_{BB}
  x_{BB}^{\Mars} +\frac{1}{2} d
w_{AB} x_{AB}^{\Mars} ) \Big),  \label{eqn.A1}
\end{equation}
\begin{equation}
x'_{AA} =\frac{1}{\bar{w} } \Big(  w_{AA} x_{AA}^{\Venus} (w_{AA} x_{AA}^{\Mars} + \frac{1}{2}
   w_{AB} x_{AB}^{\Mars}) + w_{AB} x_{AB}^{\Venus}(\frac{1}{2} w_{AA} x_{AA}^{\Mars}+ \frac{1}{2} 
  w_{AB} x_{AB}^{\Mars}) \Big),  \label{eqn.A2}
\end{equation}
and
\begin{eqnarray}
x'_{AB} = & \frac{1}{\bar{w} } \Big( w_{AA} x_{AA}^{\Venus}  ( w_{BB}  x_{BB}^{\Mars} + \frac{1}{2}
  w_{AB}  x_{AB}^{\Mars} ) +w_{BB} x_{BB}^{\Venus} ( w_{AA}  x_{AA}^{\Mars} + \frac{1}{2}
  w_{AB}  x_{AB}^{\Mars} )  \nonumber \\
& + x_{AB}^{\Venus}(\frac{1}{2}  w_{AA}  x_{AA}^{\Mars}+\frac{1}{2} (1-d)
 w_{AB}  x_{AB}^{\Mars} + \frac{1}{4}  w_{AB}
 x_{AB}^{\Mars} + (1-d)  w_{BB}  x_{BB}^{\Mars} ) \Big) \label{eqn.A3}
\end{eqnarray}
where the mean fitness of the population is
\begin{equation}
\bar{w} = w_{AA} x_{AA}+w_{AB}  x_{AB} +w_{BB}  x_{BB}.
\end{equation}
We iteration can then iterate this recursion over generations, for
example to produce the frequency trajectories shown in the inset
Figure \ref{Invasion_space} by plotting $x_{BB}+ \frac{1}{2}x_{AB}$
over time. We can also use this to assess the
invasion criteria for the allele by starting allele B at very low
frequency, and seeing if it increases across the generations.

\paragraph{Dominance} In the supplement we also explore a model where
the genotype of the male who produced the fertilizing sperm affects female drive,
instead of just the allelic type of the sperm. We assume that if a
heterozygous egg ($AB$) is fertilized by a spermwith the following
male ($\Mars$) genotypes the $B$ allele will be transmitted through
female meiosis with the follow probabilites
\begin{tabular}{cccc}
$\Mars$ geno. & AA & AB  & BB\\
B transmission prob. & $\frac{1}{2}$ &  $\frac{1}{2} + h(d-\frac{1}{2})$  & $d$\\ 
\end{tabular}
where $d > 1/2$ and $h$ ($0<h<1$) acts to specify the dominance of the $B$
allele's effect on female meiotic drive through the males genotype. We
set up a set of recursions, similar to those in eqn.
\ref{eqn.A1}-\ref{eqn.A1}, but now featuring these male genotype
transmission probabilites. In Supplementary Figure \ref{Effect_of_dominance} we show the invasion
criteria for both the A and the B allele when they are rare in the
population. We find that there is a narrow parameter range the B allele
can invade but not fix if it is dominant (i.e. $h>0$) and the cost in homozygotes
is not too great. This reflects the fact that when the allele is rare
it can increase in frequency due to fertilization of heterozygous eggs
by sperm from heterozygote males, where the sperm do not themselves
carry the allele. But as the B allele becomes common it will frequently be
present in the homozygous state in males, and so will be unable to
avoid the production of unfit homozygote zygotes leaving it trapped in
the population.

\paragraph{three allele system}
%$x_{AA}$, $x_{AB}$,$x_{AB'}$, $w_{BB'} x_{BB'}$, $w_{B'B'} x_{BB}$, $w_{B'B'} x_{B'B'}$
We denote the frequency of the three genotypes in the population at birth are
$x_{AA}$, $x_{AB}$, $x_{AB'}$, $x_{BB'}$, $x_{BB}$, $x_{B'B'}$. 
The mean relative fitness of the population is 
\begin{equation}
x_{AA}+x_{AB}+ x_{AB'}+ w_{BB'} +x_{BB'}+w_{B'B'}x_{BB} + w_{B'B'} x_{B'B'} 
\end{equation}
The frequency of the genotypes in meiotic-cells of female adults is 
\begin{equation}
\frac{1}{\bar{w}}\Big( x_{AA}, ~x_{AB},~ x_{AB'},~ w_{BB'} x_{BB'},~ w_{B'B'}x_{BB},
~w_{B'B'} x_{B'B'} \Big)
\end{equation}
The frequency of fertilization events can be found by combining these
female frequencies with the frequency of the three alleles in sperm is
\begin{eqnarray}
x_A = & x_{AA} + \frac{1}{2} x_{AB}+ \frac{1}{2} x_{AB'} \\
x_B = & w_{BB}  x_{BB} +   w_{BB'} \frac{1}{2} x_{BB'}+ \frac{1}{2} x_{AB} \\
x_{B'} = & w_{B'B'}  x_{B'B'} +   w_{BB'} \frac{1}{2} + x_{BB'}+ \frac{1}{2} x_{AB'}  
\end{eqnarray}
in combination with the transmission probabilities given in Figure \ref{Eggsperm_3_allele_cartoon}A.

\begin{table}
\begin{tabular}{lllll}
\hline 
  &  \multicolumn{4}{c}{Sperm genotype}   \\
  &  & A & B & B'  \\
$\Venus$ geno. & Freq.  & $x_{AA} + \frac{1}{2} x_{AB}+ \frac{1}{2} x_{AB'} $ & $ w_{BB}  x_{BB} +   w_{BB'} \frac{1}{2} x_{BB'}+
\frac{1}{2} x_{AB}$ & $ w_{B'B'}  x_{B'B'} +   w_{BB'} \frac{1}{2}
x_{BB'}+ \frac{1}{2} x_{AB'}  $\\
\hline\\
AA & $x_{AA}$ &  & & \\
AB & $x_{AB}$ &  & & \\
AB' & $x_{AB'}$ &  & & \\
BB' & $w_{BB'} x_{BB'}$ &  & & \\
BB & $w_{B'B'} x_{BB}$ &  & & \\
B'B' & $w_{B'B'} x_{B'B'}$ &  & & \\
\end{tabular}
\caption{ALTERNATIVELY I COULD FILL IN THIS MATRIX  with genotype }
\end{table}
\bibliography{refs}

@@

\begin{figure}
\includegraphics[width = 0.8 \textwidth]{Figures/bistable_x_vs_d_additive_s.eps} 
\caption{Bistable of a drive allele when selection cost (s) acts
  additively, for an allele whose effect
 on female meoisis is mediated by sperm allele. }  \label{bistable_additive}
\end{figure}

\begin{figure}
\includegraphics[width = 0.8 \textwidth]{Figures/effect_of_dominance_on_invasion_space_one_graph.eps} 
\caption{Invasion analysis of an allele whose effect
 on female meoisis is mediated by the genotype of the fertilizing
 male. The area below the solid/dotted line is the parameter region
 where the allele can invade/fix (from $1/1000$ and $999/1000$
 \gc{maybe should lower this}). Note
 that all of the fixation lines are on top of each other as dominance
 parameter makes little difference here. The allele will be
 polymorphic in the parameter space where the solid (invasion) line
 falls above the dotted (fixation) line. Note that this is a very
 narrow parameter range, and only exists if the allele has a
 reasonable degree on dominance}  \label{Effect_of_dominance}
\end{figure}
